---
title: "Sophia_DataExtraction"
author: "Andrew J. Temple"
date: "6/2/2024"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load packages

```{r}
## General
library(tidyverse)

## Data
library(rredlist)
#library(rfishbase)
```

Working Directory

```{r}
# Set working directory to the repository root before running
# setwd("path/to/batoid-research-bias")
```

# Data Extraction from the Red List

## Red Listings

Extract data from the RedList for all batoids

```{r}
library(iucnredlist)
api <- init_api("YOUR_IUCN_API_KEY")  # Register at https://apiv4.iucnredlist.org/ to obtain a key

chonds <- assessments_by_taxonomy(api, level = "class", name = "Chondrichthyes", latest = TRUE) %>%
  filter(scopes_description_en == "Global") %>%
  distinct()

## extract the assessments
chonds_iucn <- assessment_data_many(api, assessment_ids = chonds$assessment_id)

## filter to batoids only
unique(sapply(chonds_iucn, function(x) x$taxon$order_name))
batoids_iucn <- keep(chonds_iucn, ~ .x$taxon$order_name %in% c("RAJIFORMES", "RHINOPRISTIFORMES", "TORPEDINIFORMES", "MYLIOBATIFORMES"))

## extract species names
```


## General status (sharks vs rays)

```{r}
status_all <- data.frame()

for(i in 1:length(chonds_iucn)) {
  
temp <- data.frame(status = pluck(chonds_iucn[[i]], "red_list_category", "code", .default = NA)) %>%
  mutate(sci_name = pluck(chonds_iucn[[i]], "taxon", "scientific_name", .default = NA))

status_all <- rbind(status_all, temp) 

rm(temp)
  
}

write.csv(status_all, "status_all.csv")

## 402 of 1078 assessed species are threatened
## 229 of 561 assessed ray species are threatened
```

## Other data

Full taxonomy

```{r}
taxonomy <- data.frame()

for(i in 1:length(batoids_iucn)) {
  
temp <- data.frame(phylum = pluck(batoids_iucn[[i]], "taxon", "phylum_name", .default = NA),
                   class = pluck(batoids_iucn[[i]], "taxon", "class_name", .default = NA),
                   order = pluck(batoids_iucn[[i]], "taxon", "order_name", .default = NA),
                   family = pluck(batoids_iucn[[i]], "taxon", "family_name", .default = NA),
                   sci_name = pluck(batoids_iucn[[i]], "taxon", "scientific_name", .default = NA))

taxonomy <- rbind(taxonomy, temp) 

rm(temp)
  
}
```

Synonyms

```{r}
synonyms <- data.frame()

for(i in 1:length(batoids_iucn)) {
  
temp <- data.frame(genus = pluck(batoids_iucn[[i]], "taxon_synonyms", "genus_name", .default = NA),
                   species = pluck(batoids_iucn[[i]], "taxon_synonyms", "species_name", .default = NA)) %>%
  mutate(across(everything(), ~ str_replace_all(., " ", ""))) %>% ## strip any whitespace
  mutate(sci_name = pluck(batoids_iucn[[i]], "taxon", "scientific_name", .default = NA))

synonyms <- rbind(synonyms, temp) 

rm(temp)
  
}
```

IUCN Status

```{r}
status <- data.frame()

for(i in 1:length(batoids_iucn)) {
  
temp <- data.frame(status = pluck(batoids_iucn[[i]], "red_list_category", "code", .default = NA)) %>%
  mutate(sci_name = pluck(batoids_iucn[[i]], "taxon", "scientific_name", .default = NA))

status <- rbind(status, temp) 

rm(temp)
  
}
```

Year of description

```{r}
## to get the authority and associated year of description we also have to leverage the rredlist package
## iucnredlist doesn't contain this output for some reason
## get redlist species assessment ids
assess_id <- data.frame()

for(i in 1:length(batoids_iucn)) {
  
temp <- data.frame(id = pluck(batoids_iucn[[i]], "assessment_id", "value", .default = NA)) %>%
  mutate(sci_name = pluck(batoids_iucn[[i]], "taxon", "scientific_name", .default = NA))

assess_id <- rbind(assess_id, temp) 

rm(temp)
  
}

## get assessments through redlist and extract authority
library(rredlist)
library(purrr)

t_iucn <- map(assess_id$id, rl_assessment, key = "YOUR_IUCN_API_KEY")

authority <- data.frame()

for(i in 1:length(t_iucn)) {
  
temp <- data.frame(auth = pluck(t_iucn[[i]], "taxon", "authority", .default = NA)) %>%
  mutate(sci_name = pluck(t_iucn[[i]], "taxon", "scientific_name", .default = NA))

authority <- rbind(authority, temp) 

rm(temp)
  
}
rm(t_iucn)

## extract year of description from authority

authority <- authority %>%
  mutate(year = gsub('[[:punct:] ]+',' ', auth)) %>%
  mutate(year = parse_number(year)) %>%
  select(c(sci_name, year))
```

Depth

```{r}
depth <- data.frame()

for(i in 1:length(batoids_iucn)) {
  
temp <- data.frame(min_depth = pluck(batoids_iucn[[i]], "supplementary_info", "upper_depth_limit", .default = NA),
                   max_depth = pluck(batoids_iucn[[i]], "supplementary_info", "lower_depth_limit", .default = NA)) %>%
  mutate(sci_name = pluck(batoids_iucn[[i]], "taxon", "scientific_name", .default = NA))

depth <- rbind(depth, temp) 

rm(temp)
  
}

## first ensure depth min/maxs are the right way around (there are some errors)
depth <- depth %>%
  mutate(max_depth = if_else(max_depth >= min_depth, max_depth, min_depth),
         min_depth = if_else(min_depth <= max_depth, min_depth, max_depth))
```

Distribution (countries)

```{r}
distribution <- data.frame()

for(i in 1:length(batoids_iucn)) {
  
temp <- data.frame(origin = pluck(batoids_iucn[[i]], "locations", "origin", .default = NA),
                   description = pluck(batoids_iucn[[i]], "locations", "description", .default = NA),
                   code = pluck(batoids_iucn[[i]], "locations", "code", .default = NA)) %>%
  mutate(sci_name = pluck(batoids_iucn[[i]], "taxon", "scientific_name", .default = NA))

distribution <- rbind(distribution, temp) 

rm(temp)
  
}

## filter out non-country level locations
distribution <- distribution <- distribution %>%
  filter(!str_detect(code, "-"))

## save
#write.csv(distribution, "rl_distribution.csv", row.names = FALSE)
```


## Combine into data frame

Compile the IUCN data for analysis, save, and remove unneeded datasets

```{r}
iucn <- taxonomy %>%
  left_join(status) %>%
  left_join(depth) %>%
  left_join(authority)

## save data
#write.csv(iucn, "IUCN_Data.csv", row.names = FALSE)
```



# Other Data 

Countries of occurrence to be extracted through IUCN shapefiles
Range to be estimates from IUCN shapefiles
GDP to be extracted based on WB and countries of occurence



# Search term

create the search term from synonyms (consider com_name too)

```{r}
## drop rows where synonyms contain non-letter symbols
synonyms_df <- synonyms %>%
  filter(is.na(genus) | str_detect(genus, "^[A-Za-z]+$")) %>%
  filter(is.na(species) | str_detect(species, "^[A-Za-z]+$"))

## pull out the dropped rows, fix them and rebind to the dataf rame
synonyms_rm <- anti_join(synonyms, synonyms_df) %>%
  mutate(species = if_else(str_detect(species, "sp"), NA_character_, species),
         genus   = if_else(str_detect(species, "sp"), NA_character_, genus)) %>%
  bind_rows(data.frame(genus = c("Trygon", "Taenura", "Trygon", "Rhinobatos", "Myliobatis"),
                       species = c("henlei", "henlei", "bennettii", "leucorhinchus", "eeltenkee"),
                       sci_name = c("Potamotrygon henlei", "Potamotrygon henlei", "Hemitrygon bennetti", 
                                    "Pseudobatos leucorhynchus", "Aetobatus ocellatus"), 
                       stringsAsFactors = FALSE)) %>%
  filter(is.na(genus) | str_detect(genus, "^[A-Za-z]+$")) %>%
  filter(is.na(species) | str_detect(species, "^[A-Za-z]+$"))

synonyms_df <- rbind(synonyms_df, synonyms_rm)
rm(synonyms_rm)

## combine genus and species and add speech marks
synonyms_df <- synonyms_df %>%
  mutate(synonyms = str_c(genus, species, sep = " ")) %>%
  select(c(sci_name, synonyms)) %>%
  mutate(sci_name = str_c('"', sci_name, '"')) %>%
  mutate(synonyms = str_c('"', synonyms, '"'))

## turn them into a single string for each sci_name level, separate with OR
synonyms_df <- synonyms_df %>%
  group_by(sci_name) %>%
  summarise(synonyms = str_c(synonyms, collapse = " OR "), .groups = "drop")

## add in the scientific name to the string
synonyms_df <- synonyms_df %>%
  mutate(synonyms = str_c(sci_name, synonyms, sep = " OR ")) %>%
  mutate(synonyms = ifelse(is.na(synonyms), sci_name, synonyms))
  
## save
#write.csv(synonyms_df, "rl_synonyms.csv", row.names = FALSE)
```

Now build the terms

```{r}
## create single string 
b_search <- synonyms_df %>%
  select(synonyms) %>%
  summarise(synonyms = str_c(synonyms, collapse = " OR "), .groups = "drop") %>%
  pull()

## save
writeLines(b_search, "b_search.txt")
```