---
title: "Combining_Literature"
author: "Andrew J. Temple"
date: "3/13/2025"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

wd and libraries

```{r}
# Set working directory to the repository root before running
# setwd("path/to/batoid-research-bias")
library(tidyverse)
#library(progress)
library(doParallel)
library(foreach)
library(stringi)
```

## Load and clean literature search outputs

```{r}
# Read all files and combine them into one data frame
comb_lit <- read.csv("Literature_Search/lit_return_filtered.csv")
```

Ensure no duplicate rows

```{r}
comb_lit <- comb_lit %>%
  distinct()
```

Re-search to double check the returns and also to identify results that were based on Index Keywords - we may or may not want to filter to only those that aren't based on index keywords...

```{r}
synonyms <- read.csv("rl_synonyms.csv")

## need to change terms to the following format "a|b|c" rather than "a" OR "b" OR "c"
synonyms_check <- synonyms %>%
  mutate(terms = str_remove_all(synonyms, '"')) %>%
  mutate(terms = str_replace_all(terms, " OR ", "|"))

## unlist the make the full string
synonyms_check <- paste(unlist(synonyms_check$terms), collapse = "|")

## convert title, abstract and keyword columns lower case, also the synonym search term
## ensures the check is not case sensitive
comb_lit <- comb_lit %>%
  mutate(across(c("Title", "Abstract", "Author_Keywords", "Index_Keywords"), 
                 ~ stri_trans_general(., "Latin-ASCII"))) %>% ## deals with letters like ö
  mutate(across(all_of(c("Title", "Abstract", "Author_Keywords", "Index_Keywords")), 
                 ~ ifelse(is.na(.), NA, tolower(iconv(., from = "", to = "UTF-8")))))
## converting to UTF-8 is required to avoid some issues with tolower() and species characters

synonyms_check <- tolower(synonyms_check)


## check search terms exist in title, abstract or keywords (author and index)
## We need to run this is parallel otherwise it will take a very long time!
num_cores <- detectCores() - 2  # Use available cores minus two
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# create column for results
comb_lit$contains_terms <- logical(nrow(comb_lit))  

# Parallel loop using foreach
results <- foreach(i = 1:nrow(comb_lit), .combine = "c", .packages = "stringr") %dopar% {
  
  row_values <- unlist(comb_lit[i, c("Title", "Abstract", "Author_Keywords", "Index_Keywords")], 
                       use.names = FALSE)
  row_values <- row_values[!is.na(row_values)]  # Remove NAs
  
  any(str_detect(row_values, synonyms_check))  # Return TRUE/FALSE for each row
}

## Assign results back to the data frame
comb_lit$contains_terms <- results

## We checked the FALSE returns (n = 28)
## a quick look does suggest they do look like false returns based on our search term
## they can probably be removed

## check again to look only title, abstract or author keywords
comb_lit$contains_terms_TAK <- logical(nrow(comb_lit))  

# Parallel loop using foreach
results <- foreach(i = 1:nrow(comb_lit), .combine = "c", .packages = "stringr") %dopar% {
  
  row_values <- unlist(comb_lit[i, c("Title", "Abstract", "Author_Keywords")], use.names = FALSE)
  row_values <- row_values[!is.na(row_values)]  # Remove NAs
  
  any(str_detect(row_values, synonyms_check))  # Return TRUE/FALSE for each row
}

## Assign results back to the data frame
comb_lit$contains_terms_TAK <- results
## We checked the FALSE returns (n = 608) 

# Stop cluster
stopCluster(cl)

## remove unneeded objects
rm(cl, num_cores, results)

## save
#write.csv(comb_lit, "Literature_Search/Final_lit.csv", row.names = FALSE)
```


# Full returns including Index

## filter 

drop false returns

```{r}
filter_lit <- comb_lit %>%
  filter(contains_terms_TAK == TRUE)

## dropped 608
```

## assign number of results by species

create loop to search synonyms and count returns by species

```{r}
## make the species list
species <- synonyms %>% 
  select(sci_name) %>% 
  mutate(sci_name = str_remove_all(sci_name, '"'))

## TEST 1
num_cores <- parallel::detectCores() - 2  # Leave two cores free
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Preprocess the literature fields once
lit_fields <- tolower(
  do.call(paste, c(filter_lit[, c("Title", "Abstract", "Author_Keywords")], sep = " "))
)

start_time <- Sys.time()

# Run in parallel
counts <- foreach(i = 1:nrow(species), .combine = c, .packages = "stringr") %dopar% {
  # Build regex from synonyms (cleaned)
  terms <- gsub('"', '', synonyms[i, "synonyms"])
  terms <- tolower(gsub(" OR ", "|", terms))

  # Count matches using vectorized grepl over preprocessed text
  sum(grepl(terms, lit_fields, ignore.case = FALSE))
}

end_time <- Sys.time()
cat("Run took", round(difftime(end_time, start_time, units = "mins"), 2), "minutes\n")
stopCluster(cl)
## Time taken = 10.33 mins
## time taken 1,000 = 308.75
## time taken 10,000 = 3386.09

species <- species %>%
  mutate(counts = counts)

write.csv(species, "species_counts.csv", row.names = FALSE)

## remove unneeded objects
rm(cl, end_time, lit_fields, num_cores, start_time)
```


