---
title: "Colour"
date: "06/09/2026"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages

```{r}
library(devtools)
library(recolorize)
library(colorspace)
library(tidyverse)
library(parallel)
library(grid)
library(ggimage)
library(colordistance)
library(ape)
library(vegan)
library(cowplot)
```

Working Directory

```{r}
# Set working directory to the repository root before running
# setwd("path/to/batoid-research-bias")
```


# Image recolouration

Get images and load them

```{r}
imgs <- data.frame(file = list.files(
  "Batoid_Images/All_Images",  # Note: images not included in repo (too large/copyrighted), 
  pattern = "\\.png$", full.names = TRUE))
imgs$page <- as.numeric(str_extract(basename(imgs$file), "\\d+(?=\\.png$)"))
```

recolourise the species and check setting to see what seems most appropriate.

```{r}
recolours <- list()

recolours <- mclapply(1:nrow(imgs), function(i) {
  fit1Q <- recolorize(imgs$file[i], method = "histogram", bins = 10, plotting = FALSE, color_space = "Lab",
                      lower = c(0.95, 0.95, 0.95), upper = c(1, 1, 1))
  fit2Q <- recluster(fit1Q, cutoff = 12, plot_hclust = FALSE, plot_final = FALSE)
  fit3Q <- thresholdRecolor(fit2Q, pct = 0.01, plotting = FALSE)
  fit3Q$page <- imgs$page[i]
  return(fit3Q)
}, mc.cores = detectCores() - 2)  # Use all but two cores

rm(imgs)
```

Save recoloured images

```{r}
invisible( ## to suppress print outs
  mclapply(1:length(recolours), function(i) {
    img_array <- recoloredImage(recolours[[i]])  # e.g., a [height, width, 3] array
    
    # Convert to raster object
    img_raster <- as.raster(img_array)
    
    # Create filename
    filename <- sprintf("Batoid_Images/Recoloured/recolored_page%d.png", recolours[[i]]$page)
    
    # Save image
    png(filename, width = dim(img_array)[2], height = dim(img_array)[1], bg = "transparent")
    grid.raster(img_raster)
    dev.off()
    }, mc.cores = detectCores() - 2)  # Use all but two cores
)
```



# Image Constrast

Calculate weighted achromatic and chromatic contrast

```{r}
## Build function to calculate weighted contrast
compute_weighted_contrast <- function(contrast_matrix, weights) {
  n <- length(weights)
  if (n == 0) return(0)
  
  weighted_sum <- 0
  for (i in 1:n) {
    for (j in 1:n) {
      weighted_sum <- weighted_sum + weights[i] * weights[j] * contrast_matrix[i, j]
    }
  }
  
  return(weighted_sum)  # denominator is always 1
}


## Initialize results list
chr_contrast <- list()
achr_contrast <- list()

## build loop to run it over all species
for (i in 1:length(recolours)) {
  
  # Extract LAB cluster centers and pixel labels from fit3Q, this will still come in RGB format 
  rgb_centers <- recolours[[i]]$centers  # Each row = a color cluster in LAB
  labels <- recolours[[i]]$pixel_assignments  # Each pixel’s cluster assignment
  
  # Convert to proper Lab values
  lab_colors <- convertColor(rgb_centers, from = "sRGB", to = "Lab", scale.in = 1)

  # Get color proportions
  color_counts <- table(factor(labels, levels = seq_len(nrow(lab_colors))))
  color_props <- as.numeric(color_counts) / sum(color_counts)
  
  # Achromatic Contrast (L*)
  L_vals <- lab_colors[, 1, drop = FALSE]  # Just the L* channel
  achromatic_matrix <- as.matrix(dist(L_vals))
  
  # Chromatic Contrast (a*, b*)
  ab_vals <- lab_colors[, 2:3, drop = FALSE]  # a* and b* only
  chromatic_matrix <- as.matrix(dist(ab_vals))
  
  # Compute final scores
  weighted_achromatic <- compute_weighted_contrast(achromatic_matrix, color_props)
  weighted_chromatic <- compute_weighted_contrast(chromatic_matrix, color_props)
  
  # Store results
  chr_contrast[[i]] <- weighted_chromatic
  achr_contrast[[i]] <- weighted_achromatic
#}, mc.cores = detectCores() - 2)  # Use all but two cores
}

rm(rgb_centers, labels, lab_colors, color_counts, color_props, L_vals, chromatic_matrix, ab_vals, achromatic_matrix,
   weighted_achromatic, weighted_chromatic, i)

plot(chr_contrast, achr_contrast,
     xlab = "Chromatic contrast",
     ylab = "Achromatic contrast")
```

build data frame

```{r}
rc_img <- do.call(rbind, Map(data.frame, chr=chr_contrast, achr=achr_contrast))
rc_img$og_file <- data.frame(file = list.files(
  "Batoid_Images/All_Images"  # Note: images not included in repo (too large/copyrighted),
  pattern = "\\.png$", full.names = TRUE))
rc_img$rc_file <- list.files(
  "Batoid_Images/Recoloured", 
  pattern = "\\.png$", full.names = TRUE)
rc_img$page <- as.numeric(str_extract(basename(rc_img$rc_file), "\\d+(?=\\.png$)"))
rm(achr_contrast, chr_contrast)
```

plot in ggplot

```{r}
p01 <- ggplot(rc_img, aes(x = as.numeric(achr), y = as.numeric(chr))) +
  #geom_point() +
  geom_image(aes(image = rc_file), size = 0.05) +
  scale_x_continuous(breaks = c(0, 5, 10, 15, 20)) +
  scale_y_continuous(breaks = c(0, 5, 10, 15)) +
  xlab("Weighted Achromatic Contrast") +
  ylab("Weighted Chromatic Contrast") +
  coord_fixed() + 
  theme_bw() +
  theme(text = element_text(size = 9),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
#p01

#ggsave("contrast_12.png", p01, units = "cm", height = 12, width = 15, dpi = 600)
```

Finished with recoloured images, so remove them to save space

```{r}
rm(recolours)
```


# Colour uniqueness

create hierarchical tree of colour distance

```{r}
# create colour distance matrix
cdm <- imageClusterPipeline(rc_img$rc_file, ## match options from recolourize as much as possible
                            cluster.method = "hist", hist.bin = 5, 
                            lower = c(0.95, 0.95, 0.95), upper = c(1, 1, 1),
                            color.space = "lab", from = "sRGB", 
                            ref.white = "D65",
                            sample.size = FALSE)

# create tree and plot
tree <- nj(as.dist(cdm))
plot(tree, direction = "upwards", cex = 0.5)
```

NMDS to measure uniqueness

```{r}
## Run NDMS with different numbers of dimensions, inspect stress and select number to proceed with
stress_vals <- numeric()

for (k in 1:6) {
  set.seed(2009)
  fit <- metaMDS(as.dist(cdm), k = k, trymax = 100, trace = FALSE)
  stress_vals[k] <- fit$stress
  rm(fit)
}

plot(1:6, stress_vals, type = "b", xlab = "Dimensions (k)", ylab = "Stress", main = "Stress vs Dimensions")

# 3 dimensions takes stress to <0.05, 1 (0.229), 2 (0.075), 3 (0.035)
# Proceed with 3 dimensions

nmds_result <- metaMDS(as.dist(cdm), k = 3, trymax = 100)

# Check final stress
nmds_result$stress
# 0.035
```

Calculate relative uniqueness as distance from centroid

```{r}
# Extract NMDS site scores (coordinates)
scores_df <- as.data.frame(scores(nmds_result, display = "sites"))

# Calculate centroid
centroid <- colMeans(scores_df)

# Calculate Euclidean distance from centroid for each point
uniqueness <- apply(scores_df, 1, function(x) sqrt(sum((x - centroid)^2)))

# Add NMDS and uniqueness (distance from centroid) to data frame
rc_img <- rc_img %>%
  cbind(scores_df, uniqueness)

rownames(rc_img) <- NULL

rm(scores_df, centroid, uniqueness, cdm, nmds_result)
```

Plot colour uniqueness

```{r}
p02 <- 
  rc_img %>% 
  ggplot(aes(x = as.numeric(NMDS1), y = as.numeric(NMDS2))) +
  #geom_point() +
  geom_image(aes(image = rc_file), size = 0.05) +
  geom_text(x = -Inf, y = -Inf, label = "Stress = 0.0349",
           hjust = -0.1, vjust = -0.5, size = 9/.pt) +
  scale_x_continuous(limits = c(-40, 40), breaks = c(-40, -20, 0, 20, 40)) +
  scale_y_continuous(limits = c(-26, 26), breaks = c(-20, -10, 0, 10, 20)) +
  xlab("NMDS1") +
  ylab("NMDS2") +
  coord_fixed() +
  theme_bw() +
  theme(text = element_text(size = 10),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p02

#ggsave("uniqueness.png", p02, units = "cm", height = 10, width = 15, dpi = 600)
```



# Save Outputs

Save colour analyses data outputs

```{r}
write.csv(rc_img %>% dplyr::select(-c(og_file, rc_file)), "Colour_Data.csv", row.names = FALSE)
```

Plot combined colour plots

```{r}
p03 <- ggdraw() +
  draw_plot(plot_grid(p01, p02, ncol = 1), x = 0, y = 0, width = 1, height = 1) +
  draw_plot_label(label = c("a", "b"), size = 10,
                  x = c(0.005, 0.005), y = c(.995, 0.495))
p03

#ggsave("colour_plot.png", p03, units = "cm", height = 22, width = 15.9, dpi = 600, bg = "white")
```
