---
title: "Sophia_Analysis"
output: html_document
date: "2025-07-03"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r}
library(tidyverse)
library(dagitty)
library(ape)
library(car)
library(nnet)
library(brms)
library(tidybayes)
library(effectsize)
library(ggtree)
library(cowplot)
library(emmeans)
library(EValue)
```

Set working directory
```{r}
# Set working directory to the repository root before running
# setwd("path/to/batoid-research-bias")
```

# Data 

## build data

load data

```{r}
d <- read.csv("Data_Merged.csv")
```

check for data class

```{r}
sapply(d, class)

# change characters to factors and integers to numeric
d <- d %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(across(c("min_depth", "max_depth", "med_depth", "year", "size", "range_km2", "chr", "achr", "uniqueness", "sum_GDP_bil", "papers"), as.numeric))
```

transform highly skewed numeric variables

```{r}
hist(d$min_depth)
hist(d$max_depth)
hist(d$size)
hist(d$year)
hist(d$range_km2)
hist(d$sum_GDP_bil)
hist(d$chr)
hist(d$achr)
hist(d$uniqueness)

d <- d %>%
  mutate(min_depth = log10(min_depth+1)) %>%
  mutate(max_depth = log10(max_depth+1)) %>%
  mutate(size = log10(size+1)) %>%
  mutate(range_km2 = log10(range_km2+1)) %>%
  mutate(wsum_GDP_bil = log10(wsum_GDP_bil+1)) %>%
  mutate(sum_GDP_bil = log10(sum_GDP_bil+1)) %>%
  mutate(achr = log10(achr+1)) %>%
  mutate(chr = log10(chr+1)) %>%
  dplyr::select(c("class":"max_depth", "year", "size", "range_km2", "chr", "achr", "uniqueness":"shape", "med_lat"))
```

impute missing data

```{r}
## impute missing with mean for numeric and mode for factor
mode_value <- function(x) {
  ux <- na.omit(x)
  if (length(ux) == 0) return(NA)
  ux[which.max(tabulate(match(ux, ux)))]
}

d_imputed <- d %>%
  group_by(order) %>%
  mutate(
    across(where(is.numeric), ~ coalesce(., mean(., na.rm = TRUE))),
    across(where(is.factor), ~ {
      m <- mode_value(.)
      factor(coalesce(as.character(.), as.character(m)), levels = levels(.))
    })
  ) %>%
  ungroup()
```

Set initial reference levels for factor variables

```{r}
d_imputed$pattern <- relevel(factor(d_imputed$pattern), ref = "None")
d_imputed$shape <- relevel(factor(d_imputed$shape), ref = "Disc")
d_imputed$status <- relevel(factor(d_imputed$status), ref = "LC")
```

## Phylogeny 

Build phylogeny - in our case we will use taxonomy because a good complete phylogeny is not easily available!

```{r}
# Build tree from taxonomy
tax_tree <- as.phylo.formula(~ class/order/family/genus/sci_name, 
                             data = d)

## make tree ultametric (i.e. from class to species is equal length regardless of number of splits)
tax_tree <- compute.brlen(tax_tree, method = "Grafen")

plot(tax_tree, type = "fan", show.tip.label = FALSE)

# Create phylogenetic covariance
A <- ape::vcv(tax_tree)
```



# DAG set up

## DAG

Set up the hypothesised DAG.

```{r}
dag <- dagitty(
"dag {
contrast []
depth []
sum_GDP_bil []
papers [outcome]
pattern []
phenotype [latent]
range_km2 []
shape []
size []
status []
uniqueness []
year []
contrast -> papers
contrast -> uniqueness
depth -> papers
depth -> phenotype
depth -> range_km2
depth -> status
depth -> uniqueness
depth -> year
sum_GDP_bil -> papers
sum_GDP_bil -> status
sum_GDP_bil -> year
pattern -> papers
pattern -> uniqueness
phenotype -> contrast
phenotype -> pattern
range_km2 -> sum_GDP_bil
range_km2 -> papers
range_km2 -> status
range_km2 -> year
shape -> depth
shape -> papers
shape -> phenotype
shape -> range_km2
shape -> size
size -> depth
size -> papers
size -> range_km2
size -> status
size -> year
status -> papers
uniqueness -> papers
year -> papers
year -> status
}
")

plot(graphLayout(dag))
```


## Check for and test conditional independencies

Identify at conditional independencies. 

```{r}
iCI <- impliedConditionalIndependencies(dag)
print(impliedConditionalIndependencies(dag))
```

These should hold with the data.

Test against a correct p-value for multiple comparisons and a standardised effect size threshold of 0.05

```{r}
## use eta_squared(x, partial = TRUE) to get partial R2
#cntr _||_ rn_2 | dpth, shap
anova(lm(chr ~  
           scale(achr) + scale(min_depth) + scale(max_depth) + shape + order + family + 
           scale(range_km2),
         data = d_imputed))
anova(lm(achr ~ 
           scale(chr) + scale(min_depth) + scale(max_depth) + shape + order + family +
           scale(range_km2),
         data = d_imputed))

#cntr _||_ size | dpth, shap
anova(lm(chr ~  
           scale(achr) + scale(min_depth) + scale(max_depth) + shape + order + family +
           scale(size),
         data = d_imputed))
anova(lm(achr ~  
           scale(chr) + scale(min_depth) + scale(max_depth) + shape + order + family +
           scale(size),
         data = d_imputed))

#cntr _||_ stts | dpth, rn_2, size
anova(lm(chr ~ 
           scale(achr) + scale(min_depth) + scale(max_depth) + scale(range_km2) + scale(size) + order + family +
           status,
         data = d_imputed))
anova(lm(achr ~ 
           scale(chr) + scale(min_depth) + scale(max_depth) + scale(range_km2) + scale(size) + order + family +
           status,
         data = d_imputed))

#cntr _||_ stts | dpth, shap
anova(lm(chr ~
           scale(achr) + scale(min_depth) + scale(max_depth) + shape + order + family + 
           status,
         data = d_imputed))
anova(lm(achr ~
           scale(chr) + scale(min_depth) + scale(max_depth) + shape + order + family + 
           status,
         data = d_imputed))

#cntr _||_ w_GD | rn_2
anova(lm(chr ~  
           scale(achr) + scale(range_km2) + order + family + 
           scale(sum_GDP_bil),
         data = d_imputed))
anova(lm(achr ~  
           scale(chr) + scale(range_km2) + order + family + 
           scale(sum_GDP_bil),
         data = d_imputed))

#cntr _||_ w_GD | dpth, shap
anova(lm(chr ~ 
           scale(achr) + scale(min_depth) + scale(max_depth) + shape + order + family + 
           scale(sum_GDP_bil),
         data = d_imputed))
anova(lm(achr ~ 
           scale(chr) + scale(min_depth) + scale(max_depth) + shape + order + family + 
           scale(sum_GDP_bil),
         data = d_imputed))

#cntr _||_ year | dpth, rn_2, size
anova(lm(chr ~ 
           scale(achr) + scale(min_depth) + scale(max_depth) + scale(range_km2) + scale(size) + order + family +
           scale(year),
         data = d_imputed))
anova(lm(achr ~ 
           scale(chr) + scale(min_depth) + scale(max_depth) + scale(range_km2) + scale(size) + order + family +
           scale(year),
         data = d_imputed))

#cntr _||_ year | dpth, shap
anova(lm(chr ~
           scale(achr) + scale(min_depth) + scale(max_depth) + shape + order + family +
           scale(year),
         data = d_imputed))
anova(lm(achr ~
           scale(chr) + scale(min_depth) + scale(max_depth) + shape + order + family +
           scale(year),
         data = d_imputed))

#dpth _||_ w_GD | rn_2
anova(lm(min_depth ~  
           scale(max_depth) + scale(range_km2) + order + family + 
           scale(sum_GDP_bil),
         data = d_imputed))
anova(lm(max_depth ~  
           scale(min_depth) + scale(range_km2) + order + family + 
           scale(sum_GDP_bil),
         data = d_imputed))

#pttr _||_ rn_2 | dpth, shap
Anova(multinom(pattern ~ 
                 scale(min_depth) + scale(max_depth) + shape + order + family + 
                 scale(range_km2),
               data = d_imputed))

#pttr _||_ size | dpth, shap
Anova(multinom(pattern ~ 
                 scale(min_depth) + scale(max_depth) + shape + order + family + 
                 scale(size),
               data = d_imputed))

#pttr _||_ stts | dpth, rn_2, size
Anova(multinom(pattern ~ 
                 scale(min_depth) + scale(max_depth) + scale(range_km2) + scale(size) + order + family +
                 status,
               data = d_imputed))

#pttr _||_ stts | dpth, shap
Anova(multinom(pattern ~ 
                 scale(min_depth) + scale(max_depth) + shape + order + family +
                 status,
               data = d_imputed))

#pttr _||_ w_GD | rn_2
Anova(multinom(pattern ~ 
                 scale(range_km2) + order + family +
                 scale(sum_GDP_bil),
               data = d_imputed))

#pttr _||_ w_GD | dpth, shap
Anova(multinom(pattern ~ 
                 scale(min_depth) + scale(max_depth) + shape + order + family +
                 scale(sum_GDP_bil),
               data = d_imputed))

#pttr _||_ year | dpth, rn_2, size
Anova(multinom(pattern ~ 
                 scale(min_depth) + scale(max_depth) + scale(range_km2) + scale(size) + order + family + 
                 scale(year),
               data = d_imputed))

#pttr _||_ year | dpth, shap
Anova(multinom(pattern ~ 
                 scale(min_depth) + scale(max_depth) + shape + order + family +
                 scale(year),
               data = d_imputed))

#rn_2 _||_ unqn | cntr, dpth, pttr
anova(lm(range_km2 ~ 
           scale(chr) + scale(achr) + scale(min_depth) + scale(max_depth) + pattern + order + family +
           scale(uniqueness),
         data = d_imputed))

#rn_2 _||_ unqn | dpth, shap
summary(lm(range_km2 ~ 
             scale(min_depth) + scale(max_depth) + shape + order + family +
             scale(uniqueness),
           data = d_imputed))

#shap _||_ stts | dpth, rn_2, size
Anova(multinom(shape ~ 
                 scale(min_depth) + scale(max_depth) + scale(range_km2) + scale(size) + order + family +
                 status,
               data = d_imputed))

#shap _||_ unqn | cntr, dpth, pttr
Anova(multinom(shape ~ 
                 scale(chr) + scale(achr) + scale(min_depth) + scale(max_depth) + pattern + order + family +
                 scale(uniqueness),
               data = d_imputed))

#shap _||_ w_GD | rn_2
Anova(multinom(shape ~ 
                 scale(range_km2) + order + family +
                 scale(sum_GDP_bil),
               data = d_imputed))

#shap _||_ year | dpth, rn_2, size
Anova(multinom(shape ~ 
                 scale(min_depth) + scale(max_depth) + scale(range_km2) + scale(size) + order + family +
                 scale(year),
               data = d_imputed))

#size _||_ unqn | cntr, dpth, pttr
summary(lm(size ~ 
             scale(chr) + scale(achr) + scale(min_depth) + scale(max_depth) + pattern + order + family +
             scale(uniqueness),
           data = d_imputed))

#size _||_ unqn | dpth, shap
summary(lm(size ~ 
             scale(min_depth) + scale(max_depth) + shape + order + family + 
             scale(uniqueness),
           data = d_imputed))

#size _||_ w_GD | rn_2
summary(lm(size ~ 
             scale(range_km2) + order + family +
             scale(sum_GDP_bil),
           data = d_imputed))

#stts _||_ unqn | cntr, dpth, pttr
Anova(multinom(status ~ 
                 scale(chr) + scale(achr) + scale(min_depth) + scale(max_depth) + pattern + order + family +
                 scale(uniqueness),
               data = d_imputed))

#stts _||_ unqn | dpth, shap
Anova(multinom(status ~ 
                 scale(min_depth) + scale(max_depth) + shape + order + family +
                 scale(uniqueness),
               data = d_imputed))

#stts _||_ unqn | dpth, rn_2, size
Anova(multinom(status ~ 
                 scale(min_depth) + scale(max_depth) + scale(range_km2) + scale(size) + order + family +
                 scale(uniqueness),
               data = d_imputed))

#unqn _||_ w_GD | rn_2
summary(lm(uniqueness ~ 
             scale(range_km2) + order + family +
             scale(sum_GDP_bil),
           data = d_imputed))

#unqn _||_ w_GD | dpth, shap
summary(lm(uniqueness ~ 
             scale(min_depth) + scale(max_depth) + shape + order + family +
             scale(sum_GDP_bil),
           data = d_imputed))

#unqn _||_ w_GD | cntr, dpth, pttr
summary(lm(uniqueness ~ 
             scale(chr) + scale(achr) + scale(min_depth) + scale(max_depth) + pattern + order + family +
             scale(sum_GDP_bil),
           data = d_imputed))

#unqn _||_ year | dpth, rn_2, size
summary(lm(uniqueness ~ 
             scale(min_depth) + scale(max_depth) + scale(range_km2) + scale(size) + order + family +
             scale(year),
           data = d_imputed))

#unqn _||_ year | dpth, shap
summary(lm(uniqueness ~ 
             scale(min_depth) + scale(max_depth) + shape + order + family +
             scale(year),
           data = d_imputed))

#unqn _||_ year | cntr, dpth, pttr
summary(lm(uniqueness ~ 
             scale(chr) + scale(achr) + scale(min_depth) + scale(max_depth) + pattern + order + family +
             scale(year),
           data = d_imputed))

eta_squared(lm(uniqueness ~ 
             scale(chr) + scale(achr) + scale(min_depth) + scale(max_depth) + pattern + order + family +
             scale(year),
           data = d_imputed),
         partial = TRUE)
```

Everything holds



# Analysis!

Now conduct the analysis to identify causal relationships.

Create a data frame to hold the standardised effect sizes

```{r}
std.eff <- data.frame()
```

robustly standardise the numeric variables to facilitate approximate comparison

```{r}
## create factor to do robust scaling equivalent to standardise(x, robust = TRUE, two_sd = TRUE), this makes factor and variable effects somewhat comparable
rscale <- function(x) {
  (x - median(x, na.rm = TRUE)) / (2 * mad(x, na.rm = TRUE))
}
```

Set weakly informative priors to use for all models

```{r}
priors <- c(
  # Intercept - centred on median log1p(papers) = 1.1
  # SD of 1.5 allows range from 0 to ~4 on log1p scale
  prior(normal(1.1, 1.5), class = "Intercept"),
  
  # Coefficients - robust Student-t
  # SD of 1 allows effects up to ~1 SD of outcome
  prior(student_t(3, 0, 1), class = "b"),
  
  # Sigma - centred just below observed SD of 1.33
  # half-normal truncated at 0
  prior(normal(0, 1.5), class = "sigma")
)
```

## Accessibility

Year

```{r}
## adjustment set needed
adjustmentSets(dag, exposure = "year", outcome = "papers")

## model
m_year <- brm(
  log1p(papers) ~ rscale(year) +
    rscale(min_depth) + rscale(max_depth) + rscale(range_km2) + rscale(size) + rscale(sum_GDP_bil) +
    (1|gr(sci_name, cov = A)),   ## phylogenetic effect
  chains = 4, cores = 4, iter = 6000, warmup = 4000, threads = threading(3),
  data = d_imputed,
  data2 = list(A = A), 
  family = gaussian(),
  prior = priors,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1991
)
summary(m_year)
bayes_R2(m_year)

## calculate and bind effect size
std.eff <- rbind(std.eff, 
                 as_draws_df(m_year) %>%
                   select(starts_with("b_rscaleyear")) %>%
                   pivot_longer(everything(),
                                names_to = "Parameter",
                                values_to = "estimate") %>%
                   group_by(Parameter) %>%
                   summarise(
                     median = median(estimate),
                     lower95 = quantile(estimate, 0.025),
                     upper95 = quantile(estimate, 0.975),
                     lower50 = quantile(estimate, 0.25),
                     upper50 = quantile(estimate, 0.75)
                     ) %>%
                   mutate(Parameter = str_remove(Parameter, "b_rscale"))
)
```

Geographic Range

```{r}
## adjustment set needed
adjustmentSets(dag, exposure = "range_km2", outcome = "papers")

## model
m_range <- brm(
  log1p(papers) ~ rscale(range_km2) +
    rscale(min_depth) + rscale(max_depth) + shape + rscale(size) +
    (1|gr(sci_name, cov = A)),   ## phylogenetic effect
  chains = 4, cores = 4, iter = 6000, warmup = 4000, threads = threading(3),
  data = d_imputed,
  data2 = list(A = A),
  family = gaussian(),
  prior = priors,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1991
)
summary(m_range)
bayes_R2(m_range)

## calculate and bind effect size
std.eff <- rbind(std.eff, 
                 as_draws_df(m_range) %>%
                   select(starts_with("b_rscalerange_km2")) %>%
                   pivot_longer(everything(),
                                names_to = "Parameter",
                                values_to = "estimate") %>%
                   group_by(Parameter) %>%
                   summarise(
                     median = median(estimate),
                     lower95 = quantile(estimate, 0.025),
                     upper95 = quantile(estimate, 0.975),
                     lower50 = quantile(estimate, 0.25),
                     upper50 = quantile(estimate, 0.75)
                     ) %>%
                   mutate(Parameter = str_remove(Parameter, "b_rscale"))
)
```

GDP

```{r}
## adjustment set needed
adjustmentSets(dag, exposure = "sum_GDP_bil", outcome = "papers")

## model
m_gdp <- brm(
  log1p(papers) ~ rscale(sum_GDP_bil) +
    rscale(range_km2) + 
    (1|gr(sci_name, cov = A)),   ## phylogenetic effect
  chains = 4, cores = 4, iter = 6000, warmup = 4000, threads = threading(3),
  data = d_imputed,
  data2 = list(A = A),
  family = gaussian(),
  prior = priors,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1991
)
summary(m_gdp)
bayes_R2(m_gdp)

## calculate and bind effect size
std.eff <- rbind(std.eff, 
                 as_draws_df(m_gdp) %>%
                   select(starts_with("b_rscalesum_GDP_bil")) %>%
                   pivot_longer(everything(),
                                names_to = "Parameter",
                                values_to = "estimate") %>%
                   group_by(Parameter) %>%
                   summarise(
                     median = median(estimate),
                     lower95 = quantile(estimate, 0.025),
                     upper95 = quantile(estimate, 0.975),
                     lower50 = quantile(estimate, 0.25),
                     upper50 = quantile(estimate, 0.75)
                     ) %>%
                   mutate(Parameter = str_remove(Parameter, "b_rscale"))
)
```

Minimum & Maximum Depth

```{r}
## adjustment set needed
adjustmentSets(dag, exposure = "depth", outcome = "papers")

## model
m_depth <- brm(
  log1p(papers) ~ rscale(min_depth) + rscale(max_depth) + 
    shape + rscale(size) + 
    (1|gr(sci_name, cov = A)),   ## phylogenetic effect
  chains = 4, cores = 4, iter = 6000, warmup = 4000, threads = threading(3),
  data = d_imputed,
  data2 = list(A = A),
  family = gaussian(),
  prior = priors,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1991
)
summary(m_depth)
bayes_R2(m_depth)

## calculate and bind effect size
std.eff <- rbind(std.eff, 
                 as_draws_df(m_depth) %>%
                   select(starts_with("b_rscalemin_depth")) %>%
                   pivot_longer(everything(),
                                names_to = "Parameter",
                                values_to = "estimate") %>%
                   group_by(Parameter) %>%
                   summarise(
                     median = median(estimate),
                     lower95 = quantile(estimate, 0.025),
                     upper95 = quantile(estimate, 0.975),
                     lower50 = quantile(estimate, 0.25),
                     upper50 = quantile(estimate, 0.75)
                     ) %>%
                   mutate(Parameter = str_remove(Parameter, "b_rscale")),
                 as_draws_df(m_depth) %>%
                   select(starts_with("b_rscalemax_depth")) %>%
                   pivot_longer(everything(),
                                names_to = "Parameter",
                                values_to = "estimate") %>%
                   group_by(Parameter) %>%
                   summarise(
                     median = median(estimate),
                     lower95 = quantile(estimate, 0.025),
                     upper95 = quantile(estimate, 0.975),
                     lower50 = quantile(estimate, 0.25),
                     upper50 = quantile(estimate, 0.75)
                     ) %>%
                   mutate(Parameter = str_remove(Parameter, "b_rscale"))
)
```


## Charisma

Size

```{r}
## adjustment set needed
adjustmentSets(dag, exposure = "size", outcome = "papers")

## model
m_size <- brm(
  log1p(papers) ~ rscale(size) +
    shape +
    (1|gr(sci_name, cov = A)),   ## phylogenetic effect
  chains = 4, cores = 4, iter = 6000, warmup = 4000, threads = threading(3),
  data = d_imputed,
  data2 = list(A = A),
  family = gaussian(),
  prior = priors,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1991
)
summary(m_size)
bayes_R2(m_size)

## calculate and bind effect size
std.eff <- rbind(std.eff, 
                 as_draws_df(m_size) %>%
                   select(starts_with("b_rscalesize")) %>%
                   pivot_longer(everything(),
                                names_to = "Parameter",
                                values_to = "estimate") %>%
                   group_by(Parameter) %>%
                   summarise(
                     median = median(estimate),
                     lower95 = quantile(estimate, 0.025),
                     upper95 = quantile(estimate, 0.975),
                     lower50 = quantile(estimate, 0.25),
                     upper50 = quantile(estimate, 0.75)
                     ) %>%
                   mutate(Parameter = str_remove(Parameter, "b_rscale"))
)
```

Contrast

```{r}
## adjustment set needed
adjustmentSets(dag, exposure = "contrast", outcome = "papers")

## model
m_contrast <- brm(
  log1p(papers) ~ rscale(chr) + rscale(achr) +
    rscale(min_depth) + rscale(max_depth) + pattern + shape +
    (1|gr(sci_name, cov = A)),   ## phylogenetic effect
  chains = 4, cores = 4, iter = 6000, warmup = 4000, threads = threading(3),
  data = d_imputed,
  data2 = list(A = A),
  family = gaussian(),
  prior = priors,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1991
)
summary(m_contrast)
bayes_R2(m_contrast)

## calculate and bind effect size
std.eff <- rbind(std.eff, 
                 as_draws_df(m_contrast) %>%
                   select(starts_with("b_rscalechr")) %>%
                   pivot_longer(everything(),
                                names_to = "Parameter",
                                values_to = "estimate") %>%
                   group_by(Parameter) %>%
                   summarise(
                     median = median(estimate),
                     lower95 = quantile(estimate, 0.025),
                     upper95 = quantile(estimate, 0.975),
                     lower50 = quantile(estimate, 0.25),
                     upper50 = quantile(estimate, 0.75)
                     ) %>%
                   mutate(Parameter = str_remove(Parameter, "b_rscale")),
                 as_draws_df(m_contrast) %>%
                   select(starts_with("b_rscaleachr")) %>%
                   pivot_longer(everything(),
                                names_to = "Parameter",
                                values_to = "estimate") %>%
                   group_by(Parameter) %>%
                   summarise(
                     median = median(estimate),
                     lower95 = quantile(estimate, 0.025),
                     upper95 = quantile(estimate, 0.975),
                     lower50 = quantile(estimate, 0.25),
                     upper50 = quantile(estimate, 0.75)
                     ) %>%
                   mutate(Parameter = str_remove(Parameter, "b_rscale"))
)
```

Visual Uniqueness

```{r}
## adjustment set needed
adjustmentSets(dag, exposure = "uniqueness", outcome = "papers")

## model
m_uniq <- brm(
  log1p(papers) ~ rscale(uniqueness) +
    rscale(chr) + rscale(achr) + rscale(min_depth) + rscale(max_depth) + pattern +
    (1|gr(sci_name, cov = A)),   ## phylogenetic effect
  chains = 4, cores = 4, iter = 6000, warmup = 4000, threads = threading(3),
  data = d_imputed,
  data2 = list(A = A),
  family = gaussian(),
  prior = priors,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1991
)
summary(m_uniq)
bayes_R2(m_uniq)

## calculate and bind effect size
std.eff <- rbind(std.eff, 
                 as_draws_df(m_uniq) %>%
                   select(starts_with("b_rscaleuniqueness")) %>%
                   pivot_longer(everything(),
                                names_to = "Parameter",
                                values_to = "estimate") %>%
                   group_by(Parameter) %>%
                   summarise(
                     median = median(estimate),
                     lower95 = quantile(estimate, 0.025),
                     upper95 = quantile(estimate, 0.975),
                     lower50 = quantile(estimate, 0.25),
                     upper50 = quantile(estimate, 0.75)
                     ) %>%
                   mutate(Parameter = str_remove(Parameter, "b_rscale"))
)
```

Shape

```{r}
## adjustment set needed
adjustmentSets(dag, exposure = "shape", outcome = "papers")

## set shape order (do after seeing results)
d_imputed$shape <- factor(d_imputed$shape, levels = c("Wedge", "Disc", "Spade", "Rhombus", "Diamond"))

## model
m_shape <- brm(
  log1p(papers) ~ shape +
    (1|gr(sci_name, cov = A)),   ## phylogenetic effect
  chains = 4, cores = 4, iter = 6000, warmup = 4000, threads = threading(3),
  data = d_imputed,
  data2 = list(A = A),
  family = gaussian(),
  prior = priors,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1991
)
summary(m_shape)
bayes_R2(m_shape)

## calculate and bind effect size
std.eff_shape <- as_draws_df(m_shape) %>%
  select(b_Intercept, starts_with("b_shape")) %>%
  pivot_longer(everything(),
               names_to = "Parameter",
               values_to = "estimate") %>%
  group_by(Parameter) %>%
  summarise(
    median = median(estimate),
    lower95 = quantile(estimate, 0.025),
    upper95 = quantile(estimate, 0.975),
    lower50 = quantile(estimate, 0.25),
    upper50 = quantile(estimate, 0.75)
  ) %>%
  mutate(Parameter = str_remove(Parameter, "b_shape")) %>%
  mutate(
    lower95 = ifelse(Parameter == "b_Intercept", lower95 - median, lower95),
    upper95 = ifelse(Parameter == "b_Intercept", upper95 - median, upper95),
    lower50 = ifelse(Parameter == "b_Intercept", lower50 - median, lower50),
    upper50 = ifelse(Parameter == "b_Intercept", upper50 - median, upper50),
    median  = ifelse(Parameter == "b_Intercept", 0, median),
    Parameter = ifelse(Parameter == "b_Intercept", "Wedge", Parameter) 
  )

## pairwise comparison
## only diamond is different from reference
## use model draws for pairwise comparison
draws_s <- as_draws_df(m_shape)

draws_s <- draws_s %>%
  mutate(
    D_vs_W = b_shapeDiamond - b_Intercept,
    D_vs_D = b_shapeDiamond - b_shapeDisc,
    D_vs_S = b_shapeDiamond - b_shapeSpade,
    D_vs_R = b_shapeDiamond - b_shapeRhombus
  )

# Summarise each contrast
status_s_pairwise <- draws_s %>%
  select(D_vs_W:D_vs_R) %>%
  pivot_longer(everything(), names_to = "contrast", values_to = "estimate") %>%
  group_by(contrast) %>%
  summarise(
    median = median(estimate),
    lower95 = quantile(estimate, 0.025),
    upper95 = quantile(estimate, 0.975),
    lower50 = quantile(estimate, 0.25),
    upper50 = quantile(estimate, 0.75)
  )

rm(draws_s)
```

Pattern

```{r}
## adjustment set needed
adjustmentSets(dag, exposure = "pattern", outcome = "papers")

## set shape order (do after seeing results)
d_imputed$pattern <- factor(d_imputed$pattern, levels = c("Lined", "Other", "None", "Mottled", "Spotted"))

## model
m_pattern <- brm(
  log1p(papers) ~ pattern +
    rscale(chr) + rscale(achr) + rscale(min_depth) + rscale(max_depth) + shape +
    (1|gr(sci_name, cov = A)),   ## phylogenetic effect
  chains = 4, cores = 4, iter = 6000, warmup = 4000, threads = threading(3),
  data = d_imputed,
  data2 = list(A = A),
  family = gaussian(),
  prior = priors,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1991
)
summary(m_pattern)
bayes_R2(m_pattern)

## calculate and bind effect size
std.eff_pattern <- as_draws_df(m_pattern) %>%
  select(b_Intercept, starts_with("b_pattern")) %>%
  pivot_longer(everything(),
               names_to = "Parameter",
               values_to = "estimate") %>%
  group_by(Parameter) %>%
  summarise(
    median = median(estimate),
    lower95 = quantile(estimate, 0.025),
    upper95 = quantile(estimate, 0.975),
    lower50 = quantile(estimate, 0.25),
    upper50 = quantile(estimate, 0.75)
  ) %>%
  mutate(Parameter = str_remove(Parameter, "b_pattern")) %>%
  mutate(
    lower95 = ifelse(Parameter == "b_Intercept", lower95 - median, lower95),
    upper95 = ifelse(Parameter == "b_Intercept", upper95 - median, upper95),
    lower50 = ifelse(Parameter == "b_Intercept", lower50 - median, lower50),
    upper50 = ifelse(Parameter == "b_Intercept", upper50 - median, upper50),
    median  = ifelse(Parameter == "b_Intercept", 0, median),
    Parameter = ifelse(Parameter == "b_Intercept", "Lined", Parameter)
  )

## no need for pairwise comparison
```


## IUCN Status

IUCN Status

```{r}
## adjustment set needed
adjustmentSets(dag, exposure = "status", outcome = "papers")

## set shape order (make CR reference level because this is logical - EX will be dropped becasue n=1)
d_imputed$status <- factor(d_imputed$status, levels = c("CR", "EN", "VU", "NT", "LC", "DD", "EX"))

## model
m_status <- brm(
  log1p(papers) ~ status +
    rscale(min_depth) + rscale(max_depth) + rscale(range_km2) + rscale(size) + rscale(sum_GDP_bil) + rscale(year) + 
    (1|gr(sci_name, cov = A)),   ## phylogenetic effect
  chains = 4, cores = 4, iter = 6000, warmup = 4000, threads = threading(3),
  data = d_imputed %>% dplyr::filter(status != "EX"),
  data2 = list(A = A),
  family = gaussian(),
  prior = priors,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1991
)
summary(m_status)
bayes_R2(m_status)

## calculate and bind effect size
std.eff_status <- as_draws_df(m_status) %>%
  select(b_Intercept, starts_with("b_status")) %>%
  pivot_longer(everything(),
               names_to = "Parameter",
               values_to = "estimate") %>%
  group_by(Parameter) %>%
  summarise(
    median = median(estimate),
    lower95 = quantile(estimate, 0.025),
    upper95 = quantile(estimate, 0.975),
    lower50 = quantile(estimate, 0.25),
    upper50 = quantile(estimate, 0.75)
  ) %>%
  mutate(Parameter = str_remove(Parameter, "b_status")) %>%
  mutate(
    lower95 = ifelse(Parameter == "b_Intercept", lower95 - median, lower95),
    upper95 = ifelse(Parameter == "b_Intercept", upper95 - median, upper95),
    lower50 = ifelse(Parameter == "b_Intercept", lower50 - median, lower50),
    upper50 = ifelse(Parameter == "b_Intercept", upper50 - median, upper50),
    median  = ifelse(Parameter == "b_Intercept", 0, median),
    Parameter = ifelse(Parameter == "b_Intercept", "CR", Parameter)
  )


## evidence of difference.
## use model draws for pairwise comparisons
draws <- as_draws_df(m_status)

# CR is reference so other levels are already vs CR
draws <- draws %>%
  mutate(
    EN_vs_DD = b_statusDD - b_statusEN,
    EN_vs_LC = b_statusLC - b_statusEN,
    EN_vs_NT = b_statusNT - b_statusEN,
    EN_vs_VU = b_statusVU - b_statusEN,
    VU_vs_DD = b_statusDD - b_statusVU,
    VU_vs_LC = b_statusLC - b_statusVU,
    VU_vs_NT = b_statusNT - b_statusVU,
    NT_vs_LC = b_statusLC - b_statusNT,
    NT_vs_DD = b_statusDD - b_statusNT,
    LC_vs_DD = b_statusDD - b_statusLC

  )

# Summarise each contrast
status_pairwise <- draws %>%
  select(EN_vs_DD:LC_vs_DD) %>%
  pivot_longer(everything(), names_to = "contrast", values_to = "estimate") %>%
  group_by(contrast) %>%
  summarise(
    median = median(estimate),
    lower95 = quantile(estimate, 0.025),
    upper95 = quantile(estimate, 0.975),
    lower50 = quantile(estimate, 0.25),
    upper50 = quantile(estimate, 0.75)
  )

rm(draws)
```

IUCN Naive for comparison

```{r}
## model
m_status_naive <- brm(
  log1p(papers) ~ status + 
    (1|gr(sci_name, cov = A)),   ## phylogenetic effect
  chains = 4, cores = 4, iter = 6000, warmup = 4000, threads = threading(3),
  data = d_imputed %>% dplyr::filter(status != "EX"),
  data2 = list(A = A),
  family = gaussian(),
  prior = priors,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1991
)
summary(m_status_naive)
bayes_R2(m_status_naive)

## calculate and bind effect size
std.eff_status_naive <- as_draws_df(m_status_naive) %>%
  select(b_Intercept, starts_with("b_status")) %>%
  pivot_longer(everything(),
               names_to = "Parameter",
               values_to = "estimate") %>%
  group_by(Parameter) %>%
  summarise(
    median = median(estimate),
    lower95 = quantile(estimate, 0.025),
    upper95 = quantile(estimate, 0.975),
    lower50 = quantile(estimate, 0.25),
    upper50 = quantile(estimate, 0.75)
  ) %>%
  mutate(Parameter = str_remove(Parameter, "b_status")) %>%
  mutate(
    lower95 = ifelse(Parameter == "b_Intercept", lower95 - median, lower95),
    upper95 = ifelse(Parameter == "b_Intercept", upper95 - median, upper95),
    lower50 = ifelse(Parameter == "b_Intercept", lower50 - median, lower50),
    upper50 = ifelse(Parameter == "b_Intercept", upper50 - median, upper50),
    median  = ifelse(Parameter == "b_Intercept", 0, median),
    Parameter = ifelse(Parameter == "b_Intercept", "CR", Parameter)
  )


## evidence of difference in non-naive model.
## use model draws for pairwise comparisons.
draws_n <- as_draws_df(m_status_naive)

# CR is reference so other levels are already vs CR
draws_n <- draws_n %>%
  mutate(
    EN_vs_DD = b_statusDD - b_statusEN,
    EN_vs_LC = b_statusLC - b_statusEN,
    EN_vs_NT = b_statusNT - b_statusEN,
    EN_vs_VU = b_statusVU - b_statusEN,
    VU_vs_DD = b_statusDD - b_statusVU,
    VU_vs_LC = b_statusLC - b_statusVU,
    VU_vs_NT = b_statusNT - b_statusVU,
    NT_vs_LC = b_statusLC - b_statusNT,
    NT_vs_DD = b_statusDD - b_statusNT,
    LC_vs_DD = b_statusDD - b_statusLC
  )

# Summarise each contrast
status_n_pairwise <- draws_n %>%
  select(EN_vs_DD:LC_vs_DD) %>%
  pivot_longer(everything(), names_to = "contrast", values_to = "estimate") %>%
  group_by(contrast) %>%
  summarise(
    median = median(estimate),
    lower95 = quantile(estimate, 0.025),
    upper95 = quantile(estimate, 0.975),
    lower50 = quantile(estimate, 0.25),
    upper50 = quantile(estimate, 0.75)
  )

rm(draws_n)
```


## Phylogeny Bias

Plot bias in papers across the full phylogeny
Show a phylogenetic tree with bias in mean number of papers at each taxonomic level.

First calculate that bias

```{r}
# Log transform first
d_bias <- d_imputed %>%
  dplyr::select(c(class:sci_name, papers)) %>%
  mutate(log_papers = log1p(papers))

# Species level - raw log papers min-max scaled
d_bias <- d_bias %>%
  mutate(bias_species = (log_papers - min(log_papers)) / 
                        (max(log_papers) - min(log_papers)))

# Genus level - mean log papers per genus, then min-max scale
genus_bias <- d_bias %>%
  group_by(genus) %>%
  summarise(mean_log = mean(log_papers)) %>%
  mutate(bias_genus = (mean_log - min(mean_log)) / 
                      (max(mean_log) - min(mean_log)))

# Family level
family_bias <- d_bias %>%
  group_by(family) %>%
  summarise(mean_log = mean(log_papers)) %>%
  mutate(bias_family = (mean_log - min(mean_log)) / 
                       (max(mean_log) - min(mean_log)))

# Order level
order_bias <- d_bias %>%
  group_by(order) %>%
  summarise(mean_log = mean(log_papers)) %>%
  mutate(bias_order = (mean_log - min(mean_log)) / 
                      (max(mean_log) - min(mean_log)))

```

model phylogenetic trends in order

```{r}
## no adjustment sets needed

## model order
## set order order (do after seeing results)
d_imputed$order <- factor(d_imputed$order, levels = c("RAJIFORMES", "RHINOPRISTIFORMES", "MYLIOBATIFORMES",  "TORPEDINIFORMES"))

m_order <- brm(
  log1p(papers) ~ order, 
  chains = 4, cores = 4, iter = 6000, warmup = 4000, threads = threading(3),
  data = d_imputed,
  family = gaussian(),
  prior = priors,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1991
)
summary(m_order)
bayes_R2(m_order)

## calculate and bind effect size
std.eff_order <- as_draws_df(m_order) %>%
  select(b_Intercept, starts_with("b_order")) %>%
  pivot_longer(everything(),
               names_to = "Parameter",
               values_to = "estimate") %>%
  group_by(Parameter) %>%
  summarise(
    median = median(estimate),
    lower95 = quantile(estimate, 0.025),
    upper95 = quantile(estimate, 0.975),
    lower50 = quantile(estimate, 0.25),
    upper50 = quantile(estimate, 0.75)
  ) %>%
  mutate(
    lower95 = ifelse(Parameter == "b_Intercept", lower95 - median, lower95),
    upper95 = ifelse(Parameter == "b_Intercept", upper95 - median, upper95),
    lower50 = ifelse(Parameter == "b_Intercept", lower50 - median, lower50),
    upper50 = ifelse(Parameter == "b_Intercept", upper50 - median, upper50),
    median  = ifelse(Parameter == "b_Intercept", 0, median),
    Parameter = ifelse(Parameter == "b_Intercept", "b_orderRAJIFORMES", Parameter)
  ) %>%
  mutate(Parameter = str_remove(Parameter, "b_order")) 
```

model phylogenetic trends in family

```{r}
## no adjustment sets needed

## model family
## set family order (do after seeing results) - set the lowest rajiform first, then keep in low-high within order (for later visualisation)
d_imputed$family <- factor(d_imputed$family, 
                           levels = c(
                             ## Rajiformes
                             "RAJIDAE", "ANACANTHOBATIDAE", "GURGESIELLIDAE", "ARHYNCHOBATIDAE", 
                             ## Rhinopristiformes
                             "RHINOBATIDAE", "RHINIDAE", "TRYGONORRHINIDAE", "GLAUCOSTEGIDAE", "PRISTIDAE",
                             ## Myliobatiformes
                             "UROLOPHIDAE", "ZANOBATIDAE", "HEXATRYGONIDAE", "PLESIOBATIDAE", 
                             "POTAMOTRYGONIDAE", "AETOBATIDAE", "UROTRYGONIDAE", "MYLIOBATIDAE", 
                             "GYMNURIDAE", "DASYATIDAE", "RHINOPTERIDAE", "MOBULIDAE", 
                             ## Torpediniformes
                             "NARCINIDAE", "NARKIDAE", "HYPNIDAE", "PLATYRHINIDAE", "TORPEDINIDAE"))

m_family <- brm(
  log1p(papers) ~ family, 
  chains = 4, cores = 4, iter = 6000, warmup = 4000, threads = threading(3),
  data = d_imputed,
  family = gaussian(),
  prior = priors,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1991
)
summary(m_family)
bayes_R2(m_family)

## calculate and bind effect size
std.eff_family <- as_draws_df(m_family) %>%
  select(b_Intercept, starts_with("b_family")) %>%
  pivot_longer(everything(),
               names_to = "Parameter",
               values_to = "estimate") %>%
  group_by(Parameter) %>%
  summarise(
    median = median(estimate),
    lower95 = quantile(estimate, 0.025),
    upper95 = quantile(estimate, 0.975),
    lower50 = quantile(estimate, 0.25),
    upper50 = quantile(estimate, 0.75)
  ) %>%
  mutate(
    lower95 = ifelse(Parameter == "b_Intercept", lower95 - median, lower95),
    upper95 = ifelse(Parameter == "b_Intercept", upper95 - median, upper95),
    lower50 = ifelse(Parameter == "b_Intercept", lower50 - median, lower50),
    upper50 = ifelse(Parameter == "b_Intercept", upper50 - median, upper50),
    median  = ifelse(Parameter == "b_Intercept", 0, median),
    Parameter = ifelse(Parameter == "b_Intercept", "b_familyRAJIDAE", Parameter) 
  ) %>%
  mutate(Parameter = str_remove(Parameter, "b_family")) 
```



# Finalise Plotting

## Phylogenetic bias plotting

Phylo effects (tree)

```{r}
tax_bias <- bind_rows(
  order_bias  %>% transmute(label = order,    bias = bias_order,  level = "Order",   lwd = 4),
  family_bias %>% transmute(label = family,   bias = bias_family, level = "Family",  lwd = 3),
  genus_bias  %>% transmute(label = genus,    bias = bias_genus,  level = "Genus",   lwd = 2),
  d_bias      %>% transmute(label = sci_name, bias = bias_species,        level = "Species", lwd = 1)
)

## Build a clean taxonomy tree from data
## Every species will be exactly the same depth from root

tax_paths <- d_bias %>%
  distinct(class, order, family, genus, sci_name) %>%
  mutate(path = paste(class, order, family, genus, sci_name, sep = "/")) %>%
  pull(path)

tree_clean <- as.phylo(~ class/order/family/genus/sci_name, 
                        data = distinct(d_bias, class, order, family, genus, sci_name),
                        collapse = FALSE)

tree_clean <- compute.brlen(tree_clean, 1)

p_species <- ggtree(tree_clean, layout = "circular", colour = NA) %<+% tax_bias +
  geom_tree(aes(colour = bias, linewidth = lwd)) +
  geom_tiplab(size = 1/.pt, hjust = -0.1) +
  scale_colour_distiller(
    palette  = "Reds",
    direction = 1,        # 1 = light to dark, -1 = dark to light
    na.value = "grey60",
    name     = "Research Bias\nWithin\nTaxonomic Level\n(scaled log(Papers+1))",
    guide     = guide_colourbar(title.hjust = 0.5)
  ) +
  scale_linewidth(range = c(0.25, 1.5), guide = "none") +  # tune min/max thickness here
  theme_tree() +
  theme(text = element_text(size = 8))

# clean
rm(tax_bias, tax_paths, tree_clean)
```

plot for order

```{r}
p_eff_order <- 
  ggplot(std.eff_order %>%
           mutate(across(c(median, lower95, upper95, lower50, upper50),
                         ~ if_else(Parameter == "RAJIFORMES", 0, .))),
         aes(x = reorder(Parameter, median), y = median)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), linewidth = .5, alpha = 1, fatten = 1) +
  geom_pointrange(aes(ymin = lower50, ymax = upper50), linewidth = 1, alpha = 1, fatten = 1) +
  geom_point(size = 1) +
  scale_size_manual(values = c("TRUE" = 2, "FALSE" = 1), guide = "none") +
  geom_hline(yintercept = 0, linetype = 2, alpha = 1, colour = "grey50") +
  geom_vline(xintercept = 1.5, linetype = 1, alpha = .5, colour = "grey50") +
  geom_vline(xintercept = 2.5, linetype = 1, alpha = .5, colour = "grey50") +
  geom_vline(xintercept = 3.5, linetype = 1, alpha = .5, colour = "grey50") +
  scale_x_discrete(position = "top", labels = c("Rajiformes", "Torpediniformes", "Myliobatiformes", "Rhinopristiformes"),
                   expand = c(0.15, 0)) +
  scale_y_continuous(limits = c(-0.1,NA)) +
  labs(y = "Standardised Effect Size", x = "Order") +
  theme_bw() +
  theme(text = element_text(size = 8),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

plot for family

```{r}
## other than Rajidae, order within order as least to most researched.
std.eff_family$Parameter <- factor(std.eff_family$Parameter, 
                           levels = c(
                             ## Rajiformes
                             "RAJIDAE", "GURGESIELLIDAE", "ANACANTHOBATIDAE", "ARHYNCHOBATIDAE", 
                             ## Torpediniformes
                             "NARKIDAE", "HYPNIDAE", "NARCINIDAE", "PLATYRHINIDAE", "TORPEDINIDAE",
                             ## Myliobatiformes
                             "UROLOPHIDAE", "HEXATRYGONIDAE", "ZANOBATIDAE", "PLESIOBATIDAE", 
                             "POTAMOTRYGONIDAE", "UROTRYGONIDAE", "DASYATIDAE", "GYMNURIDAE",
                             "MYLIOBATIDAE", "AETOBATIDAE", "RHINOPTERIDAE", "MOBULIDAE",
                             ## Rhinopristiformes
                             "RHINOBATIDAE", "RHINIDAE", "GLAUCOSTEGIDAE", "TRYGONORRHINIDAE", "PRISTIDAE"
                             )
                           )

p_eff_family <- 
  ggplot(std.eff_family %>% 
           mutate(across(c(median, lower95, upper95, lower50, upper50),
                         ~ if_else(Parameter == "RAJIDAE", 0, .))) %>%
           mutate(Parameter = factor(str_to_sentence(Parameter), 
                                     levels = str_to_sentence(levels(Parameter)))), aes(x = Parameter, y = median)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), linewidth = .5, alpha = 1, fatten = 1) +
  geom_pointrange(aes(ymin = lower50, ymax = upper50), linewidth = 1, alpha = 1, fatten = 1) +
  geom_point(size = 1) +
  scale_size_manual(values = c("TRUE" = 2, "FALSE" = 1), guide = "none") +
  geom_hline(yintercept = 0, linetype = 2, alpha = 1, colour = "grey50") +
  geom_vline(xintercept = 4.5, linetype = 1, alpha = .5, colour = "grey50") +
  geom_vline(xintercept = 9.5, linetype = 1, alpha = .5, colour = "grey50") +
  geom_vline(xintercept = 21.5, linetype = 1, alpha = .5, colour = "grey50") +
  scale_x_discrete(expand = c(0.025, 0)) +
  labs(y = "Standardised Effect Size", x = "Family") +
  theme_bw() +
  theme(text = element_text(size = 8),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

plot together

```{r}
p_tax_bias <- 
  ggdraw() +
  draw_plot(p_species, 
            x = 0, y = .5, width = 1, height = .5) +
  draw_plot(plot_grid(p_eff_order + theme(axis.title.y = element_blank()),
                      p_eff_family + theme(axis.title.y = element_blank()),
                      ncol = 1, align = "v", rel_heights = c(.5, 1)),
            x = 0.03, y = 0, width = .97, height = .5) +
  draw_label("Standardised Effect Size", 
             size = 8, angle = 90, x = 0.01, y = .29, vjust = 1) +
  draw_plot_label(label = c("a", "b"), size = 8,
                  x = c(0.01, .01), y = c(.99, .49)) +
  draw_line(x = c(0.226, 0.305), y = c(0.3205, 0.3462), colour = "grey50", alpha = .5) +
  draw_line(x = c(0.398, 0.535), y = c(0.3205, 0.3462), colour = "grey50", alpha = .5) +
  draw_line(x = c(0.809, 0.764), y = c(0.3205, 0.3462), colour = "grey50", alpha = .5)

#ggsave("Taxa_Bias.jpeg", p_tax_bias, width = 12, height = 16, units = "cm", dpi = 600, bg = "white")
```


## Standardised effect sizes for downstream causal variables

plot standardised effect sizes for all causal numeric variables

```{r}
p_eff <- ggplot(std.eff, aes(x = median, y = Parameter, )) +
  geom_pointrange(aes(xmin = lower95, xmax = upper95), linewidth = .5, alpha = 1, fatten = 1) +
  geom_pointrange(aes(xmin = lower50, xmax = upper50), linewidth = 1, alpha = 1, fatten = 1) +
  geom_point(size = 1) +
  geom_vline(xintercept = 0, linetype = 2, alpha = 1, colour = "grey50") +
  scale_y_discrete(labels = c("Achromatic\nContrast", "Chromatic\nContrast", "Maximum\nDepth", "Minimum\nDepth",
                              "Geographic\nRange", "Maximum\nSize", "GDP", "Colour\nUniqueness", "Year of\nDiscovery")) +
  scale_x_continuous(limits = c(-2.8, 2.8)) +
  labs(x = "Standardised Effect Size", title = "Numeric Drivers") +
  theme_bw() +
  theme(text = element_text(size = 8),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5, size = 8))
```

plot standardised effect sizes for shape

```{r}
## Wedge is reference level
p_eff_shape <- ggplot(std.eff_shape %>%
                        mutate(across(c(median, lower95, upper95, lower50, upper50),
                         ~ if_else(Parameter == "Wedge", 0, .))),
           aes(x = median, y = reorder(Parameter, -median))) +
  geom_pointrange(aes(xmin = lower95, xmax = upper95), linewidth = .5, alpha = 1, fatten = 1) +
  geom_pointrange(aes(xmin = lower50, xmax = upper50), linewidth = 1, alpha = 1, fatten = 1) +
  geom_point(size = 1) +
  scale_size_manual(values = c("TRUE" = 2, "FALSE" = 1), guide = "none") +
  geom_vline(xintercept = 0, linetype = 2, alpha = 1, colour = "grey50") +
  scale_y_discrete(position = "right") +
  scale_x_continuous(limits = c(-2.8, 2.8)) +
  labs(x = "Standardised Effect Size", title = "Body Shape") +
  theme_bw() +
  theme(text = element_text(size = 8),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5, size = 8))
```

plot standardised effect sizes for pattern

```{r}
## Lined is reference level
p_eff_pattern <- ggplot(std.eff_pattern %>%
                          mutate(across(c(median, lower95, upper95, lower50, upper50),
                         ~ if_else(Parameter == "Lined", 0, .))),
           aes(x = median, y = reorder(Parameter, -median))) +
  geom_pointrange(aes(xmin = lower95, xmax = upper95), linewidth = .5, alpha = 1, fatten = 1) +
  geom_pointrange(aes(xmin = lower50, xmax = upper50), linewidth = 1, alpha = 1, fatten = 1) +
  geom_point(size = 1) +
  scale_size_manual(values = c("TRUE" = 2, "FALSE" = 1), guide = "none") +
  geom_vline(xintercept = 0, linetype = 2, alpha = 1, colour = "grey50") +
  scale_y_discrete(position = "right") +
  scale_x_continuous(limits = c(-2.8, 2.8)) +
  labs(x = "Standardised Effect Size", title = "Patterning") +
  theme_bw() +
  theme(text = element_text(size = 8),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5, size = 8))
```

plot standardised effect sizes for status

```{r}
std.eff_status <- std.eff_status %>%
  mutate(Parameter = fct_relevel(Parameter, "CR", "EN", "VU", "NT", "LC", "DD"))

## CR is reference level
p_eff_status <- ggplot(std.eff_status %>%
                         mutate(across(c(median, lower95, upper95, lower50, upper50),
                         ~ if_else(Parameter == "CR", 0, .))),
           aes(x = median, y = Parameter)) +
  geom_pointrange(aes(xmin = lower95, xmax = upper95, colour = Parameter), linewidth = .5, alpha = 1, fatten = 1) +
  geom_pointrange(aes(xmin = lower50, xmax = upper50, colour = Parameter), linewidth = 1, alpha = 1, fatten = 1) +
  geom_point(size = 1, aes(colour = Parameter), shape = 21, fill = "black") +
  scale_colour_manual(values = c("#D81E05", "#FC7F3F", "#F9E814", "#CCE226", "#60C659", "#D1D1C6")) +
  geom_vline(xintercept = 0, linetype = 2, alpha = 1, colour = "grey50") +
  scale_y_discrete(position = "right") +
  scale_x_continuous(limits = c(-2.8, 2.8)) +
  labs(x = "Standardised Effect Size", title = "IUCN Status") +
  theme_bw() +
  theme(text = element_text(size = 8),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.title.y = element_blank(), 
        legend.position = "none", 
        plot.title = element_text(hjust = 0.5, size = 8))
p_eff_status
```

Plot together

```{r}
p_causal <- ggdraw() +
  draw_plot(p_eff, 
            x = 0, y = 0, width = .5, height = 1) +
  draw_plot(plot_grid(p_eff_shape, p_eff_pattern, p_eff_status,
                      ncol = 1, rel_heights = c(5, 5, 6), align = "v"), 
            x = .5, y = 0, width = .5, height = 1) 

#ggsave("Causal_Drivers.jpeg", p_causal, width = 12, height = 12, units = "cm", dpi = 600, bg = "white")
```



# Sensitivity testing

Test sensitivity of the causal relationship to modeled confounders. Sensitivity using e-values is on a risk ratio scale. larger values = lower risk. Risk is subject to unmeasured confounders, high risk doesn't mean unreliable model outcomes.

```{r}
## Function to allow looping across models
## Takes a brms model and the name of the causal predictor.
## Returns the E-value table from evalues.OLS().
evalue_brms <- function(model, predictor) {
  fe    <- fixef(model, summary = TRUE)
  beta  <- fe[predictor, "Estimate"]
  lower <- fe[predictor, "Q2.5"]
  upper <- fe[predictor, "Q97.5"]
  sigma <- summary(model)$spec_pars["sigma", "Estimate"]

  d  <- beta / sigma
  se <- ((upper - lower) / sigma) / (2 * 1.96)

  evalues.OLS(est = d, se = se, sd = 1)
}

## Apply across models 
model_specs <- list(
  ## access
  list(model = m_year, predictor = "rscaleyear"),
  list(model = m_range, predictor = "rscalerange_km2"),
  list(model = m_gdp, predictor = "rscalesum_GDP_bil"),
  list(model = m_depth, predictor = "rscalemin_depth"),
  list(model = m_depth, predictor = "rscalemax_depth"),
  ## charisma
  list(model = m_size, predictor = "rscalesize"),
  list(model = m_contrast, predictor = "rscalechr"),
  list(model = m_contrast, predictor = "rscalechr"),
  list(model = m_uniq, predictor = "rscaleuniqueness"),
  list(model = m_shape, predictor = "shapeDisc"),
  list(model = m_shape, predictor = "shapeSpade"),
  list(model = m_shape, predictor = "shapeRhombus"),
  list(model = m_shape, predictor = "shapeDiamond"), 
  list(model = m_pattern, predictor = "patternOther"),
  list(model = m_pattern, predictor = "patternNone"), 
  list(model = m_pattern, predictor = "patternMottled"), 
  list(model = m_pattern, predictor = "patternSpotted"),
  ## status
  list(model = m_status, predictor = "statusEN"),
  list(model = m_status, predictor = "statusVU"),
  list(model = m_status, predictor = "statusNT"),
  list(model = m_status, predictor = "statusLC"),
  list(model = m_status, predictor = "statusDD")
  # add as needed
)

evalue_results <- lapply(model_specs, function(x) {
  ev <- evalue_brms(x$model, x$predictor)
  data.frame(
    predictor = x$predictor,
    evalue_point = ev["E-values", "point"],
    evalue_ci    = ev["E-values", "lower"]
  )
})

evalue_table <- do.call(rbind, evalue_results)
evalue_table
```
